---
title: "Lab 8: Mini Project"
author: "Tusha (A17339806)"
format: pdf
toc: true
---


## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class. You’ll extend what you’ve learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses. This expands on our RNA-Seq analysis from last day.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.


## Data import

Data was downloaded as a csv file.

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
head(wisc.df)
```

Remove the first diagnosis column since we want to use unsupervised methods to classify this data.

```{r}
wisc.data <- wisc.df[,-1]
head(wisc.data)
```

Saving the diagnosis column so we can refer to it later.

```{r}
diagnosis <- wisc.df[,1]
```

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

There are `r nrow(wisc.data)` observations/patients in this dataset.

> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(diagnosis == "M")
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.data))) # returns column numbers of the column names with "_mean"
grep("_mean", colnames(wisc.data), value=TRUE) # returns column names with "_mean"
```

## Principal Component Analysis

Need to do EDA to check if the data is scaled.

```{r}
colMeans(wisc.data)
```

```{r}
apply(wisc.data,2,sd)
```

### Perform PCA

Then we go ahead and perform PCA. The `prcomp()` function has a `scale=FALSE` default. We nearly always want to set this as `TRUE` since we don't want our analysis to be domnicated by columns/variables in our dataset that have high standard deviation and mean when compared to others just because the measurements are on different units/scales.

```{r}
# scaling makes the sd=1
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

About 44.27% of the original variance is captured by PC1.

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7

### Visualize PCA results

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

The plot is awful.

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

Since PC2 explains more variance than PC3, the groups are more separated in the first plot

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- wisc.pr$sdev^2 / sum(wisc.pr$sdev^2)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
wisc.pr$rotation[,1]
```


> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

-0.26085376

> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

5


## Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

data.dist <- dist(data.scaled)

wisc.hclust <- hclust(data.dist, method="complete")
wisc.hclust
```

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```

For 4 clusters, we need to cut it off at ~19.

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, h=19)
table(wisc.hclust.clusters, diagnosis)
```

For the above, Group 1 seems to be malignant patients and Group 3 seems to be benign patients.

> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

I think either 4 or 8 groups seem to be the best choices for cluster vs diagnoses matches.

```{r}
wisc.hclust.sample <- hclust(data.dist, method="complete")
plot(wisc.hclust.sample)
wisc.hclust.clusters.sample <- cutree(wisc.hclust.sample, k=8)
table(wisc.hclust.clusters.sample, diagnosis)
```

```{r}
wisc.hclust.sample1 <- hclust(data.dist, method="single")
plot(wisc.hclust.sample1)
wisc.hclust.clusters.sample1 <- cutree(wisc.hclust.sample1, k=4)
table(wisc.hclust.clusters.sample1, diagnosis)
```

```{r}
wisc.hclust.sample2 <- hclust(data.dist, method="complete")
plot(wisc.hclust.sample2)
wisc.hclust.clusters.sample2 <- cutree(wisc.hclust.sample2, k=4)
table(wisc.hclust.clusters.sample2, diagnosis)
```

```{r}
wisc.hclust.sample3 <- hclust(data.dist, method="average")
plot(wisc.hclust.sample3)
wisc.hclust.clusters.sample3 <- cutree(wisc.hclust.sample3, k=4)
table(wisc.hclust.clusters.sample3, diagnosis)
```

```{r}
wisc.hclust.sample4 <- hclust(data.dist, method="ward.D2")
plot(wisc.hclust.sample4)
wisc.hclust.clusters.sample4 <- cutree(wisc.hclust.sample4, k=4)
table(wisc.hclust.clusters.sample4, diagnosis)
```



> Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

The `ward.D2` option does seem the best since it has the most clear distinction between the B and M groups. That is, it breaks the group into clusters that have the most clear 


## Combining methods

### Clustering based on PCA results

Clustering the original data was not very productive. The PCA results looked promising. So clustering in PC space will help.

```{r}
# take the first 3 PCs 
dist.pc <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(dist.pc, method="ward.D2")
```

```{r}
plot(wisc.pr.hclust)
abline(h=70, col="red")
```

To get our clustering membership vector, we cut the tree.

```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```

> Q15. How well does the newly created model with four clusters separate out the two diagnoses?

How does this clustering compare to the actual diagnosis?

```{r}
table(grps, diagnosis)
```

> Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters, diagnosis)
```

## Specificity and Sensitivity

Sensitivity refers to a test’s ability to correctly detect ill patients who do have the condition. In our example here the sensitivity is the total number of samples in the cluster identified as predominantly malignant (cancerous) divided by the total number of known malignant samples. In other words: TP/(TP+FN).

Specificity relates to a test’s ability to correctly reject healthy patients without a condition. In our example specificity is the proportion of benign (not cancerous) samples in the cluster identified as predominantly benign that are known to be benign. In other words: TN/(TN+FN).


## Predictions

```{r}
url <- "new_samples.csv"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
g <- as.factor(grps)
levels(g)
g <- relevel(g,2)
levels(g)
```


```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q18. Which of these new patients should we prioritize for follow up based on your results?

We should prioritize the patients in Group 1.