---
title: "Lab 6: Gradebook"
author: "Tusha (A17339806)"
format: pdf
---

The below is my automated gradebook project for lab06:

```{r}
# Reading in the csv file for the gradebook
data <- read.csv(file="student_homework.csv", row.names = 1)

# Other gradebook vectors for testing
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```


***Question 1***

The grade function:

```{r}
grade <- function(scores)
{
  # Making a copy of the vector to modify
  scores_for_grading <- scores
  
  # is.na() creates a logical vector for scores in the input vector that are missing
  # We then replace those scores with -1 (lower than any possible score) to make sure that it is not included in our calculations
  scores_for_grading[is.na(scores_for_grading)] <- -1
  
  # Check which score is the lowest
  lowest_index <- which.min(scores_for_grading)
  scores_without_lowest <- scores_for_grading[-lowest_index]
  
  # If there are any unresolved NAs, we remove them
  scores_without_lowest <- scores_without_lowest[scores_without_lowest != -1]
  
  # Compute the mean
  mean(scores_without_lowest)
}
```

Testing:

```{r}
student3_grade <- grade(student3)
print(student3_grade)
```

***Question 2***

To find out who has the top scoring student in the class is, we have to calculate average scores for everyone and determine who has the highest one.

```{r}
# This applies the grade function to each row of the gradebook, i.e. once for every student, and saves the returned scores as a vector
final_grades <- apply(data, 1, grade)
print(final_grades)
```
```{r}
top_scorer_index <- which.max(final_grades)
print(top_scorer_index)
```

Therefore, the top scoring student in the class is Student 18.

***Question 3***

To find out which homework was the toughest, we have to calculate average scores for each and determine which led to students performing the worst.

```{r}
# This applies the grade function to each column of the gradebook, i.e. once for every homework, and saves the returned scores as a vector
final_grades_per_hw <- apply(data, 2, grade)
print(final_grades_per_hw)
```
```{r}
lowest_score_index <- which.min(final_grades_per_hw)
print(lowest_score_index)
```

Therefore, the toughest homework was Homework 2.

