---
title: "Lab 7: Machine Learning 1"
author: "Tusha (A17339806)"
format: pdf
---

Fundamentals of machine learning including clustering and dimensionality reduction

Creating a dataset:

```{r}
hist(rnorm(500,mean=5))
```

```{r}
x <- c(rnorm(30,mean=-3),rnorm(30,mean=3))
y <- rev(x)

data <- cbind(x,y)
```

## K-means clustering

```{r}
k <- kmeans(data, 2)
k
```


> Question: How many points are in each cluster?

```{r}
k$size
```

> What components of your result object details:
  - cluster assignment/membership?
  - cluster center?
  
```{r}
k$cluster
k$centers
```

> Make a clustering results figure of the data colored by cluster membership

```{r}
plot(data, col=c("red", "blue"))
```

```{r}
plot(data, col=k$cluster)
points(k$centers, col="royalblue", pch=20, cex=2)

# pch is plotting character i.e. dot type
# cex is character expansion
```

K-means clustering is very popular as it is very fast and relatively straightforward: it takes numeric data as input and return the cluster membership, etc.

The problem is that we tell k-means how many clusters we want

> Run kmeans again and cluster it into 4 groups and plot the results like above

```{r}
k4 <- kmeans(data, 4)
plot(data, col=k4$cluster)
points(k4$centers, col="royalblue", pch=20, cex=2)
```

It picks lowest value of total variation within clusters. That is the tightest fit and returns that.
There is a large reduction in SS at k=2 but after that the values do not go down as quickly.
 
```{r}
k1 <- kmeans(data, 1)
k3 <- kmeans(data, 3)
k5 <- kmeans(data, 5)

ss <- c(k1$tot.withinss, k$tot.withinss, k3$tot.withinss, k4$tot.withinss, k5$tot.withinss)
ss
```

```{r}
plot(ss, type="b", xlab="k", ylab="total within ss")
```

```{r}
n <- NULL

for (i in 1:5)
{
  n <- c(n, kmeans(data, i)$tot.withinss)
}

plot(n, type="b")
```

## Hierarchichal Clustering

We can't just input our data. We need to first calculate a distance matrix for our data (e.g. `dist()`) and use it as input for the `hclust()` function.
Not just euclidean distance but also sequence similarity, amino acid similarity, etc.

```{r}
d <- dist(data) 
# 60x60 matrix with the distance of every point in the dataset with every other point

hc <- hclust(d)
hc
```

There is a plot method for hclust results

```{r}
plot(hc)
abline(h=8, col="red")
```

The numbers between 1-30 and 31-60 are on separate sides of the graphs i.e. they are highly separated by distance.

To get our cluster membership vector (i.e. our main clustering result) we can "cut" the tree at a given height or a height that yields a given "k" number of groups.

```{r}
hccol <- cutree(hc, h=8)
```

> Plot the data with our hclust result coloring

```{r}
plot(data, col=hccol)
```

## Principal Component Analysis

### PCA of UK food data

```{r}
url <- "https://tinyurl.com/UK-foods"
food <- read.csv(url, row.names=1)
dim(food)
head(food)
```

```{r}
# rownames(food) <- food [,1]
# food <- food[,-1]
# food

# Can get disruptive if run again (keeps removing columns)
```

Some base figures

```{r}
barplot(as.matrix(food), beside=T, col=rainbow(nrow(food)))
```

```{r}
barplot(as.matrix(food), beside=F, col=rainbow(nrow(food)))
```

There is one plot that can be useful for small datasets

```{r}
pairs(food, col=rainbow(10), pch=16)
```


> It can be challengin to spot major trends and patters even in relatively small multivariate datasets (here we have 17, typically we have 1000s)


### PCA

In base R, the main PCA function is `prcomp()`

I will take the transpose of our food data so the "foods" are in the columns

```{r}
pca <- prcomp(t(food))
summary(pca)
```

```{r}
pca$x
```

```{r}
cols <- c("black","red","blue","darkgreen")
plot(pca$x[,1], pca$x[,2], col=cols, pch=16, xlab="PCA1", ylab="PCA2")
```

```{r}
library(ggplot2)
```

```{r}
ggplot(pca$x) +
  aes(PC1, PC2) +
  geom_point(col=cols) +
  theme_minimal()
```

```{r}
pca$rotation

# How much each of these variables weighs on each of the axes/pcs
# Telling us what makes two countries different from each other (based on PCA results)
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col() +
  theme_minimal()
```

